Information-theoretic measures for non-linear causality detection: \\ application to social media sentiment and cryptocurrency prices

### 2020-07-22 paper/submission.tex
Change Price->Returns in plots, and fix table references (hardcoded)

  In order to validate the autoregressive and information-theoretic approaches to detecting causality, we apply these to the calculation of transfer entropy for synthetic data generated by both linear and non-linear coupled time series, of increasing coupling strength. 
  The computations are performed over the relative returns of the synthetic signals generated as described in sections \ref{s.method}\ref{s.LinearSignal} and \ref{s.method}\ref{s.NonLinearSignal}. 
  It is important to note that many areas for which such statistical studies are important are constrained by limited available data. Therefore, for a technique of causal inference to be useful, it must perform even with small sample sizes. In Section \ref{s.results} we apply both classical and information-theoretic methodologies to financial data with sample sizes of the order of $10^4$, and so we first evaluate the performance using synthetic time series of $15,000$ observations, with known causal coupling, before confirming that the technique can detect the same causal coupling with as few as $500$ samples. The results of this evaluation for increasing signal strength, using sample sizes of $500$, $1000$, $2500$, $10,000$ and $100,000$, are presented in Table \ref{t.table2} in the appendix.

@@

  \begin{figure*}[!htbp]
    \includegraphics[width=\linewidth]{images/BTC_6_quantile_bins.pdf}
    \caption{Evidence that BTC sentiment and price are causally coupled in both directions in a non-linear way. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, show a high level of significance, especially during 2017 and 2018, in both directions, with the peak significance $Z=21.3$ observed in the direction of price to sentiment in the 24-month window ending February 2018.}
    \label{fig:BTC_TE}

    \includegraphics[width=\linewidth]{images/LTC_6_quantile_bins.pdf} 
    \caption{Evidence that LTC price and sentiment are causally coupled in both directions in a non-linear way, with sentiment having a larger influence on price than the other way round. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, show a significant signal in both directions, with the net information transfer generally operating in the direction of sentiment to price. Peak significance of $Z=42.8$ is observed from sentiment to price in the 24-month window ending February 2018.}
    \label{fig:LTC_TE}
    \vspace{-4pt}
  \end{figure*}

  \begin{figure*}[!htbp]
    \includegraphics[width=\linewidth]{images/XRP_6_quantile_bins.pdf}
    \caption{Evidence that XRP price and sentiment are causally coupled in both directions in a non-linear way, with the prevailing direction of information transfer flowing from sentiment to price in the first period, and from price to sentiment in the second. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, show a small but clear significant signal, in both directions, which decays rapidly towards January 2018 and does not recover afterward. Peak significance of $Z=41.9$ is observed from sentiment to price in the 24-month window ending January 2018.}
    \label{fig:XRP_TE}

    \includegraphics[width=\linewidth]{images/ETH_6_quantile_bins.pdf}
    \caption{Evidence that ETH price and sentiment are causally coupled in both directions in a non-linear way. Overall this coupling is of lower significance compared to the other currencies investigated. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, indicate initial significance for windows ending before 2018, followed by low significance after the collapse in signal strength, which must therefore begin around January 2016. Peak significance is observed in the first window of the study, with $Z=20.6$ in the direction of sentiment to price, and $Z=19.9$ in the direction of price to sentiment.     
  }
  \label{fig:ETH_TE}
  \vspace{-34pt}
  \end{figure*}

@@

The significance tests confirm the existence of causally-coupled relationships, though the strength of these relationships are challenging to accurately quantify from the data, especially for the purposes of comparison between different time series, and between the linear and non-linear results over the same data. However, the significance values themselves offer the possibility of quantifying the strength of causality, which may be used as a proxy when using transfer entropy as a tool for detecting statistical causality. With this work we demonstrate that the dynamics of the causative relationship is non-linear, as the autoregressive technique was able to detect only small or zero causality in either direction, for any of the currencies. The mean Z-scores of results observed by linear and information-theoretic methods are presented for each currency in Table \ref{t.table4}. Our results show that histograms of marginal equiquantisation, in concert with permutation tests, can be used in practice to detect nonlinear information transfer between time series, even where classical statistical techniques fail. Let us point out that there is a risk of assuming ergodicity in the results; we have shown that the level of causation varies historically within the sample time period, and there is no evidence that the observed causal relationships will continue unchanged.

@@

  The social sentiment data was provided courtesy of PsychSignal, and may be made available pending request to the authors. The data takes the form of the number of positive messages and the number of negative messages, publicly shared on either Twitter or StockTwits, associated each hour with the cryptocurrencies in question. The association is detected via the use of a `hashtag' (or `cashtag') which takes the form of \#BTC or \#Bitcoin (for example) on twitter, or \$BTC on StockTwits.  For inclusion in the dataset, the message must contain one of the tags described in Table \ref{tab:table1}.

@@

  Analysis was undertaken to validate the performance of linear and non-linear techniques over multiple parameter values for sample size and signal strength. Table \ref{t.table2} contains the Z-scores of transfer entropy calculations using marginal equiquantisation, for coupled random walks of increasing coupling strength and sample size.

  Table \ref{t.table3} contains results quantifying the net information transfer across windows of the study for each cryptocurrency. Net information transfer is calculated as $ \operatorname{TE}^{(1)}_{X \rightarrow Y} - \operatorname{TE}^{(1)}_{Y \rightarrow X} $, and is given in units of bits.

@@

  \begin{table*}[!htb]
    \label{t.table3}
    \caption{ Net information transfer statistics in units of bits determined over all windows of the study for each cryptocurrency. Mean values show that in time, more information was transfered from sentiment to price for LTC and XRP, with more information transfer in the direction of price to sentiment for BTC and ETH. It should be noted that mean due to the overlapping strides of the windows, periods during the middle of the sample are weighted more heavily; however the figures appropriately present the relative direction of information transfer over the 48-month period of the study, so are presented for the sake of comparison. Significance scores are presented in Table \ref{t.table4}}

>>>>

  In order to validate the autoregressive and information-theoretic approaches to detecting causality, we apply these to the calculation of transfer entropy for synthetic data generated by both linear and non-linear coupled time series, of increasing coupling strength. 
  The computations are performed over the relative returns of the synthetic signals generated as described in sections \ref{s.method}\ref{s.LinearSignal} and \ref{s.method}\ref{s.NonLinearSignal}. 
  It is important to note that many areas for which such statistical studies are important are constrained by limited available data. Therefore, for a technique of causal inference to be useful, it must perform even with small sample sizes. In Section \ref{s.results} we apply both classical and information-theoretic methodologies to financial data with sample sizes of the order of $10^4$, and so we first evaluate the performance using synthetic time series of $15,000$ observations, with known causal coupling, before confirming that the technique can detect the same causal coupling with as few as $500$ samples. The results of this evaluation for increasing signal strength, using sample sizes of $500$, $1000$, $2500$, $10,000$ and $100,000$, are presented in Table {\color{blue}2} in the appendix.

@@

  \begin{figure*}[!htbp]
    \includegraphics[width=\linewidth]{images/BTC.pdf}
    \caption{Evidence that BTC sentiment and price are causally coupled in both directions in a non-linear way. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, show a high level of significance, especially during 2017 and 2018, in both directions, with the peak significance $Z=21.3$ observed in the direction of price to sentiment in the 24-month window ending February 2018.}
    \label{fig:BTC_TE}

    \includegraphics[width=\linewidth]{images/LTC.pdf} 
    \caption{Evidence that LTC price and sentiment are causally coupled in both directions in a non-linear way, with sentiment having a larger influence on price than the other way round. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, show a significant signal in both directions, with the net information transfer generally operating in the direction of sentiment to price. Peak significance of $Z=42.8$ is observed from sentiment to price in the 24-month window ending February 2018.}
    \label{fig:LTC_TE}
    \vspace{-4pt}
  \end{figure*}

  \begin{figure*}[!htbp]
    \includegraphics[width=\linewidth]{images/XRP.pdf}
    \caption{Evidence that XRP price and sentiment are causally coupled in both directions in a non-linear way, with the prevailing direction of information transfer flowing from sentiment to price in the first period, and from price to sentiment in the second. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, show a small but clear significant signal, in both directions, which decays rapidly towards January 2018 and does not recover afterward. Peak significance of $Z=41.9$ is observed from sentiment to price in the 24-month window ending January 2018.}
    \label{fig:XRP_TE}

    \includegraphics[width=\linewidth]{images/ETH.pdf} 
    \caption{Evidence that ETH price and sentiment are causally coupled in both directions in a non-linear way. Overall this coupling is of lower significance compared to the other currencies investigated. Non-linear TE is calculated by multidimensional histograms with 6 quantile bins per dimension. Z-scores, calculated over 100 shuffles, indicate initial significance for windows ending before 2018, followed by low significance after the collapse in signal strength, which must therefore begin around January 2016. Peak significance is observed in the first window of the study, with $Z=20.6$ in the direction of sentiment to price, and $Z=19.9$ in the direction of price to sentiment.     
  }
  \label{fig:ETH_TE}
  \vspace{-34pt}
  \end{figure*}

@@

The significance tests confirm the existence of causally-coupled relationships, though the strength of these relationships are challenging to accurately quantify from the data, especially for the purposes of comparison between different time series, and between the linear and non-linear results over the same data. However, the significance values themselves offer the possibility of quantifying the strength of causality, which may be used as a proxy when using transfer entropy as a tool for detecting statistical causality. With this work we demonstrate that the dynamics of the causative relationship is non-linear, as the autoregressive technique was able to detect only small or zero causality in either direction, for any of the currencies. The mean Z-scores of results observed by linear and information-theoretic methods are presented for each currency in Table {\color{blue}4}. Our results show that histograms of marginal equiquantisation, in concert with permutation tests, can be used in practice to detect nonlinear information transfer between time series, even where classical statistical techniques fail. Let us point out that there is a risk of assuming ergodicity in the results; we have shown that the level of causation varies historically within the sample time period, and there is no evidence that the observed causal relationships will continue unchanged.

@@

  The social sentiment data was provided courtesy of PsychSignal, and may be made available pending request to the authors. The data takes the form of the number of positive messages and the number of negative messages, publicly shared on either Twitter or StockTwits, associated each hour with the cryptocurrencies in question. The association is detected via the use of a `hashtag' (or `cashtag') which takes the form of \#BTC or \#Bitcoin (for example) on twitter, or \$BTC on StockTwits.  For inclusion in the dataset, the message must contain one of the tags described in Table {\color{blue}1}.

@@

  Analysis was undertaken to validate the performance of linear and non-linear techniques over multiple parameter values for sample size and signal strength. Table {\color{blue}2} contains the Z-scores of transfer entropy calculations using marginal equiquantisation, for coupled random walks of increasing coupling strength and sample size.

  Table {\color{blue}3} contains results quantifying the net information transfer across windows of the study for each cryptocurrency. Net information transfer is calculated as $ \operatorname{TE}^{(1)}_{X \rightarrow Y} - \operatorname{TE}^{(1)}_{Y \rightarrow X} $, and is given in units of bits.

@@

  \begin{table*}[!htb]
    \label{t.table3}
    \caption{ Net information transfer statistics in units of bits determined over all windows of the study for each cryptocurrency. Mean values show that in time, more information was transfered from sentiment to price for LTC and XRP, with more information transfer in the direction of price to sentiment for BTC and ETH. It should be noted that mean due to the overlapping strides of the windows, periods during the middle of the sample are weighted more heavily; however the figures appropriately present the relative direction of information transfer over the 48-month period of the study, so are presented for the sake of comparison. Significance scores are presented in Table {\color{blue}4}.}
